Environmental pathogen reservoirs and habitat heterogeneity in a metapopulation

### 2014-04-21 Manuscript/manuscript.tex
Discussion edits

To understand the mechanisms behind these effects, we further examined within patch dynamics on high variance metapopulations that produced endemic dynamics.  Because of their lower extinction rate and higher recolonization rate, high quality patches support more consistent occupancy than low quality patches (Fig~\ref{simvis}).  However, as an epidemic progresses, high quality patches increasingly support infected occupants, while low quality patches support infecteds infrequently (Fig~\ref{simvis}).  In addition, while high quality patches do support susceptible occupancy, they experience a greater number of infection events (susceptible to infected transitions) than low quality patches (Fig~\ref{infections}).  This relationship becomes weaker as longevity increases, due to the fact that as longevity increases, there are fewer susceptibles remaining in the metapopulation to become infected.

@@

The positive relationship between pathogen persistence/spread and environmental longevity demonstrated in the theoretical model above is well documented in models of empirical systems.  \cite{Almberg2011} find that, in plausible simulations of chronic wasting disease in deer, increasing prion survival in the environment increases the resulting force of infection and peak prevalence, with the  importance of reservoir transmission increasing through time.  Similarly, Breban et al 2009 show that environmental transmission of avian influenza virus allows it to persist in situations where direct transmission alone is insufficient.  \emph{Other examples.}

When longevity is sufficient for the pathogen to persist, we find that habitat heterogeneity controls how and where it spreads.  In particular, increased habitat quality variance facilitates pandemic dynamics, increasing the likelihood that the pathogen will spread through the entire host population.  The mechanisms underlying this behavior can best be understood by drawing parallels between our metapopulation model and models of pathogen spread through a contact network.  In this framework, our metapopulation becomes a network with patches as nodes connected through immigration (with the strength of the connection measured by connectivity).  Many studies have shown that networks with high degree variance (where a node's degree is roughly equivalent to its connectivity in our metapopulation model) are easily invaded by pathogens, which are then able to spread much more rapidly than in less heterogeneous networks (\cite{Pastor-Satorras2001}).

This increase in pathogen spread in high variance metapopulations is largely attributable to the presence of high quality patches (Figure 2).  Since metapopulation connectivity is determined entirely by patch quality (recall $d_{ij}=1$ for all $i, j$), high quality patches function as highly connected hubs.  Borrowing vocabulary from network theory, high quality patches are effectively high degree nodes (i.e. many connections).  Specifically, the assumptions of the model mean that high quality patches have both high in-degree (attract colonists from other patches) and a high out-degree (produce colonists that spread to other patches).  These two properties have important implications for pathogen spread in the metapopulation. The high in-degree helps position high quality patches as the ecological "traps" hypothesized above, wherein susceptibles repeatedly colonize high quality patches and subsequently become infected (Figure 4).  This infection pressure is the result of both the environmental reservoir and direct infection from infected immigrants attracted to the high quality patch.  This reflects the results from the contact network literature that show that highly connected (i.e. high degree) nodes have higher infection risk than less connected nodes (\cite{Christley2005}, \cite{Keeling2005}).  This trap effect, coupled with the relatively low extinction rate on high quality patches helps to create a stable platform from which the pathogen can spread through the rest of the metapopulation.  Indeed, high quality patches are effectively metapopulation scale superspreaders (\cite{Lloyd-Smith2005}).  Once the pathogen infects high quality patches, it is able to spread to the rest of the metapopulation relatively easily.  This process helps to maintain infected patches throughout the metapopulation, which in turn feed back on high quality patches to ensure a steady stream of infected colonists that maintain the trap effect.  Thus, through these two interacting mechanisms -- the trap and superspreader effects -- the presence of high quality patches in high variance metapopulations serves to significantly increase pathogen spread.

Despite the dominating effects of high quality patches, low quality patches still play an important role in some cases by providing refuges on which susceptibles can escape infection. Even though individual low quality patches are unable to support consistent occupancy (susceptible or otherwise), at the metapopulation level, the presence of low quality patches increases overall susceptible occupancy, suggesting that in aggregate, low quality patches can provide susceptible refuges.  This phenomenon represents the other side of the hub role played by high quality patches in that low quality patches are relatively weakly connected to the metapopulation and are therefore infrequently colonized by infected individuals.  This is consistent with the result that low degree nodes tend to be the last infected in models of pathogen spread on contact networks (\cite{Barthelemy2004}).  Moreover, the instability of low quality patches (i.e. the high extinction rate), together with the increased mortality of infected populations, means that infected individuals do not persist long on low quality patches.   However, it is important to caveat this refuge effect by noting that it appears to be swamped in the presence of too many high quality patches, as seen by the fact that the likelihood of pandemic increases with patch quality variance despite the fact that high variance metapopulations have equal proportions of high and low quality patches.  This suggests that once enough high quality patches are added, the force of infection in the system becomes large enough to overwhelm the weak connectivity of low quality patches.

>>>>

To understand the mechanisms behind these effects, we further examined within patch dynamics in high variance metapopulations that produced endemic dynamics.  Because of their lower extinction rate and higher recolonization rate, high quality patches support more consistent occupancy than low quality patches (Fig~\ref{simvis}).  However, as an epidemic progresses, high quality patches increasingly support infected occupants, while low quality patches support infecteds infrequently (Fig~\ref{simvis}).  In addition, while high quality patches do support susceptible occupancy, they experience a greater number of infection events (susceptible to infected transitions) than low quality patches (Fig~\ref{infections}).  This relationship becomes weaker as longevity increases, due to the fact that as longevity increases, there are fewer susceptibles remaining in the metapopulation to become infected.

@@

The positive relationship between pathogen spread and environmental longevity demonstrated in the theoretical model above is well documented in models of empirical systems.  \cite{Almberg2011} find that, in plausible simulations of chronic wasting disease in deer, increasing prion survival in the environment increases the resulting force of infection and peak prevalence, with the  importance of reservoir transmission increasing through time.  Similarly, Breban et al 2009 show that environmental transmission of avian influenza virus allows it to persist in situations where direct transmission alone is insufficient.  \emph{Other examples.}

Moreover, for a given longevity, these simulations suggest that high variance metapopulations are more prone to widespread pathogen spread than low variance metapopulations.  The mechanisms underlying this behavior can be understood by drawing parallels with models of pathogen spread through a contact network.  In this context, our metapopulation can be viewed a network with patches as nodes connected through immigration (with the strength of the connection measured by connectivity).  Many studies have shown that networks with high degree variance (where a node's degree is roughly equivalent to its total connectivity in our metapopulation model) are easily invaded by pathogens, which are then able to spread much more rapidly than in less heterogeneous networks (\cite{Pastor-Satorras2001}).

In addition to allowing us to explore the effects of longevity and variance independently, this experiment also allowed us to vary the two simultaneously and explore their interaction, suggesting that the specific effect of longevity on pathogen spread is mediated by the variance of the patch quality distribution.  In particular, metapopulations with high patch quality variance require lower environmental longevities to experience pandemic dynamics.  In this way, the effects of environmental longevity are exacerbated by a high variance patch quality distribution, especially for intermediate longevities.

This interaction between environmental longevity and the variance of the patch quality distribution is largely attributable to the presence of high quality patches.  Since a patch's connectivity is determined, at least in part, by its quality, high quality patches function as highly connected hubs in the metapopulation.  Specifically, high quality patches have both high in-degree (attract colonists from other patches) and a high out-degree (produce colonists that spread to other patches).  These two properties have important implications for pathogen spread in the metapopulation. The high in-degree helps position high quality patches as the ecological "traps" hypothesized above, wherein susceptible hosts repeatedly colonize high quality patches and subsequently become infected (Fig~\ref{infections}).  This infection pressure results in part from direct transmission from infected immigrants attracted to the high quality patch, reflecting results from the contact network literature that show that highly connected nodes have higher infection risk (\cite{Christley2005}, \cite{Keeling2005}), but is further augemented by environmental transmission.

This trap effect, coupled with the relatively low extinction rate on high quality patches helps to create a stable platform from which the pathogen can spread through the rest of the metapopulation.  Indeed, high quality patches are effectively metapopulation scale superspreaders (\cite{Lloyd-Smith2005}).  Once the pathogen infects high quality patches, it is able to spread to the rest of the metapopulation relatively easily.  This process helps to maintain infected patches throughout the metapopulation, which in turn feed back on high quality patches to ensure a steady stream of infected colonists that maintain the trap effect.  Thus, through these two interacting mechanisms -- the trap and superspreader effects -- the presence of high quality patches serves to significantly increase pathogen spread.

Despite the dominating effects of high quality patches, low quality patches still play an important role in some cases by providing refuges on which susceptibles can escape infection. Even though individual low quality patches are unable to support consistent occupancy, at the metapopulation level, the presence of low quality patches increases overall susceptible occupancy, an effect that is strongest at intermediate longevities.   This phenomenon represents the other side of the hub role played by high quality patches in that low quality patches are relatively weakly connected to the metapopulation and are therefore infrequently colonized by infected individuals.  Moreover, the instability of low quality patches (i.e. the high extinction rate), together with the increased mortality of infected populations, means that infected populations do not persist long on low quality patches.  Because low quality patches are so infrequently colonized, the environmental reservoir left behind by these infected populations generally decays before it has the opportunity to infect new susceptible colonists.

Due to their lower extinction rates, high quality patches are able to maintain stable occupancy despite widespread infection.  On the other hand, at high pathogen longevities, the positive effect of low qualtiy distributions on susceptible occupancy is overwhelmed, and low quality patches are unable to support infected occupancy, resulting in an increase in host extinction events.  Thus, the role of the patch quality distribution and the presence of high or low quality patches is generally greatest for low to intermediate pathogen longevity -- once longevity is high enough, the pathogen persists and spreads easily enough that the effects of individual patches are washed out.  Then the role of patch quality lies primarily in its effect on extinction rates, and whether infected hosts can persist.
